Optimal updating factor in Wang-Landau and metadynamics simulations

### 2016-04-14 doc/invt.tex
invt.tex: edit to Sec. II.C.3

Free energy calculation\cite{frenkel, newman} is a central theme
in computational physics and chemistry.
%
Given a system,
the task is to compute,
via either Monte Carlo\cite{frenkel, newman, landau_binder} (MC)
and molecular dynamics\cite{frenkel, karplus2002} (MD) simulations,
a distribution, $p^*(z)$,
along a quantity of interest $z$.
%
The negative logarithm,
$-\log p^*(z)$,
defines a free energy,
or the potential of mean force (PMF).
%
Normally, the distribution $p^*(z)$ is
limited in one or a few narrow $z$ regions.
%
Therefore, to capture the global shape
of the PMF,
one has to run multiple independent simulations
under different umbrella or bias potentials
along $z$\cite{mezei1987, berg1992, lee1993},
and then piece together the entire curve.
%
This solution can be laborious % clumsy in practice
and it often fails to converge quickly
for a complex and glassy system,
because the system can be trapped
in a local free energy minimum.
%for a long time.

@@

We develop the theory
in the following order.
%
In Sec. \ref{sec:background},
we first review the basics of
flat-distribution sampling
and some known aspects on the optimal schedule.
%
This also helps fixing the notations.
%
Then, in Sec. \ref{sec:single-bin},
we illustrate the method of
computing the optimal schedule
on the simplest case of
single-bin (WL) updating scheme.
%
We shall prove the optimality
of the known inverse-time formula.
%
Next, we turn to the general case
of multiple-bin updating schemes
in Secs. \ref{sec:multiple-bin}
to \ref{sec:band-matrix}.
%
Finally, we compare different updating schemes
in the long time limit
in Sec. \ref{sec:cmpschemes}.

@@

Given a system,
consider the problem of computing
the distribution, $p^*_i$,
along a discrete quantity $i$.
%
%The discreteness of $i$ is natural
%in discrete models.
%
For example, $i$ can be the energy $E$
in a lattice spin model; or the temperature index
in a simulated tempering simulation.
%
For a continuous quantity, $z$,
which is typical in molecular systems,
we can discretize $z$
such that the integer $i$ represents
the index of a small interval, or a bin,
$(z, z + dz)$.
%
In both cases,
the distribution is normalized as
$\sum_{i = 1}^n p^*_i = 1$.

For a large system,
the distribution $p^*_i$ tends to
be localized,
%
and to find out the global shape,
it is often advantageous to carry out
a biased sampling that targets
a wider distribution $p_i$.
%
%Here, we refer to simulations that target
%a flat or nearly flat distribution
%as entropic or multicanonical sampling.

To do so, we need to introduce a bias potential $V_i$
to modify the target distribution to
%
\begin{equation}
  \pi_i \propto p^*_i \, e^{-V_i}.
  \label{eq:pi_p_phi1}
\end{equation}
%
Upon normalization, $\sum_{i = 1}^n \pi_i = 1$,
we get
%
\begin{equation}
  \pi_i
  =
  \frac{                p^*_i \, e^{-V_i} }
       { \sum_{j = 1}^n p^*_j \, e^{-V_j} }
  .
  \label{eq:pi_p_phi}
\end{equation}
%
Particularly,
to achieve a flat distribution $\pi_i$,
the bias potential $V_i$
must coincide with $\log p^*_i$,
up to an additive constant.

Here we shall call a simulation that targets
a nearly flat distribution\cite{dayal2004, trebst2004, singh2011},
$p_i$,
a flat-distribution sampling (FDS) simulation.
%
Roughly speaking, there are two types of FDS methods.
%
The first type,
referred to as the equilibrium FDS, % (E-FDS) methods hereinafter,
estimates the bias potential, $V_i$,
before simulation,
and fixes it during simulation.
%
Examples include the original
entropic or multicanonical sampling\cite{berg1992, lee1993}
and simulated tempering\cite{marinari1992, lyubartsev1992}
with $i$ being the index of energy
and that of temperature, respectively.
%
The estimated bias potential, $V_i$,
are often improved
by the histogram, $H_i$, along $i$ accumulated
during simulation.
%
Using $H_i$ as $\pi_i$ in
Eq. \eqref{eq:pi_p_phi1}
allows us to correct $V_i$ as
%
\begin{equation}
V^\mathrm{corr}_i
=
V_i
+
\log \frac{ H_i }
          { p_i }.
\label{eq:vcorr_equil}
\end{equation}
%
\note{This follows from
  $$
  \begin{aligned}
    \pi_i &\propto p^*_i \, e^{-V_i}, \\
    p_i &\propto p^*_i \, e^{-V^\mathrm{corr}_i}.
  \end{aligned}
  $$
}
For an unknown system,
the main difficulty of the process
is that it is hard to guess a sufficiently accurate
$V_i$ to start with.
%
Thus,
it can take many iterative rounds of simulations
to achieve the desired flat distribution.
%
Further, a reasonably precise histogram
requires an accumulation period
longer than the autocorrelation time,
which can also be hard to estimate beforehand.

For this reason, the second type of methods,
referred to as the adaptive FDS methods,
were invented.
%
Examples include
the WL and metadynamics methods.
%
These methods update the bias potential more frequently
and can easily achieve a flat distribution.
%
However, they are less well understood theoretically,
as the updating loops used there
break the microscopic reversibility
of the underlying sampling mechanism.
%
%Thus, for a conservative practitioner,
%the E-FDS methods appear more appealing
%once a sufficiently accurate bias potential is available.
%
To minimize this effect,
many adaptive methods gradually
reduce the updating magnitude
over the course of simulation\cite{marsili2006,
liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008}.
%
Thus, in late stages of a long simulation,
the updating is negligible and
such an adaptive simulation is rather similar
to the corresponding equilibrium simulation.
%
If the bias potential were converged to the exact value,
this particular equilibrium simulation can be thought as
the optimal asymptotic limit of the adaptive counterpart,
delivering the maximal efficiency.

For later convenience, we introduce a shifted bias potential
%
\begin{equation}
  v_i \equiv V_i - \ln p^*_i + \ln p_i.
  \label{eq:v_def}
\end{equation}
%
Since the last two terms of Eq. \eqref{eq:v_def}
remain constant during the course of simulation,
updates to $V_i$'s are equivalent to those to $v_i$'s.
%
In terms of $v_i$'s, Eq. \eqref{eq:pi_p_phi}
becomes
%
\begin{equation}
  \pi_i
  =
  \frac{                p_i \, e^{-v_i} }
       { \sum_{j = 1}^n p_j \, e^{-v_j} }
  \propto
  p_i \, e^{-v_i}.
  \label{eq:pi_p_v}
\end{equation}
%
According to Eq. \eqref{eq:pi_p_phi1},
$v_i$ should approach a constant of $i$
upon convergence (i.e., when $\pi_i \approx p_i$)
of an adaptive FDS.

@@

In the WL algorithm\cite{wang2001, wang2001pre},
the bias potential, $v_i(t)$, is updated
in each MC step $t$ as
%
\begin{equation}
  v_i(t+1)
  =
  v_i(t)
  +
  \delta_{i, \, i(t)}
  \frac{ \alpha(t) } { p_i }.
  \label{eq:wl_update}
\end{equation}
%
where $i(t)$ is the bin at step $t$,
and $\alpha(t)$ is the updating magnitude.
%
We refer to the above scheme as the \emph{single-bin scheme},
for it applies only to the bias potential
at the current bin, $i(t)$.
%
In a more general \emph{multiple-bin scheme}, however,
several neighboring bins are updated:
%
\begin{equation}
  v_i(t+1)
  =
  v_i(t)
  +
  w_{i, \, i(t)}
  \frac{ \alpha(t) } { p_i },
  \label{eq:mbin_update}
\end{equation}
%
where $w_{ij}$ are a set of coefficients satisfying
$\sum_{j=1}^n w_{ij} = 1$.
%
We shall discuss other restrictions on $w_{ij}$
in Sec. \ref{sec:updating-matrix}.
%
The Gaussian updating scheme
used by metadynamics
is a multiple-bin scheme.

One can show that with a constant $\alpha(t) = \alpha > 0$,
the distribution collected from
the trajectory is identical to $p_i$ in long times.
%
However, as the above updates
change $v_i(t)$ continuously,
Eq. \eqref{eq:pi_p_v} no longer holds
at all times.
%
In other words,
the error in $v_i(t)$
depends on the updating magnitude\cite{
  zhou2005, liang2005, laio2005, bussi2006, poulain2006, liang2007,
  morozov2007, zhou2008, morozov2009, crespo2010,
  atchade2011, fort2015}.
%%
%The source of the deviation is two-fold.
%%
%First, since $v_i(t)$ is updated continuously,
%there is a random noise that is proportional
%to $\sqrt \alpha$.
%%
%Second, there is a systematic error
%that comes from the updating dynamics itself,
%as it breaks the Markovian nature
%that underlies Eq. \eqref{eq:pi_p_v}.
%%
%
Thus, one has to decrease $\alpha$ over time
to reduce the error of $v_i$,
in order to get the unbiased distribution
$\ln p_i^*$ via Eq. \eqref{eq:v_def}.

In the original WL scheme,
the updating magnitude $\alpha$ (or $\ln f$
in the original paper) is kept as a constant
for a period of time,
referred to as a stage below.
%
The histogram collected in the stage is monitored.
%
Once the histogram is sufficiently flat,
we are allowed to enter a new stage
with $\alpha$
reduced by a factor of one
half\cite{wang2001, wang2001pre}.
%
This scheme works well for early stages.
%
However, it tends to reduce $\alpha$
too quickly in late stages, making the asymptotic error
saturate\cite{belardinelli2007, belardinelli2007jcp, belardinelli2008}.
%

@@

%For the single-bin scheme
%used in WL algorithm,
Later,
it was found that
a better way
of updating the schedule $\alpha(t)$
in late stages
is to follow the formula
%
\begin{equation}
  \alpha(t) = \frac{1}{t},
  \label{eq:alpha_invt}
\end{equation}
%
where $t$ is the number of steps
from the beginning of the MC simulation,
which shall be referred to as the ``time'' below.
%
This simple yet effective formula has attracted
several studies\cite{belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008, morozov2009,
komura2012, caparica2012, caparica2014}.
%
There are, however, some reservation about
the constant of proportionality:
the optimal formula for $\alpha(t)$
might be $C/t$ with a constant $C$
different from $1.0$\cite{
morozov2007, zhou2008, morozov2009}.

As Eq. \eqref{eq:alpha_invt} was being developed for the WL algorithm,
several similar ways of
reducing the updating magnitude
were developed for umbrella sampling and metadynamics,
particularly the
self-healing umbrella sampling
and well-tempered metadynamics\cite{
  marsili2006, barducci2008, dickson2011, dama2014}.
%
Our aim below is to the optimal schedule $\alpha(t)$,
for a general updating scheme.

\subsection{\label{sec:single-bin}
Single-bin scheme}

@@

We first approximate Eq. \eqref{eq:wl_update}
by a differential equation
%
\begin{equation}
  \dot v_i(t)
  =
  \alpha(t) \, \frac{ h_i(t) } { p_i },
  \label{eq:vt_diffeq}
\end{equation}
%
where
$\dot v_i(t) \equiv dv_i(t)/dt$,
%
and
\begin{equation}
  h_i(t) = \delta_{i, i(t)}
  ,
  \label{eq:h_def}
\end{equation}
is the instantaneous histogram,
which is equal to $1.0$
for the current bin $i(t)$
or zero otherwise.

Next, we split $h_i(t)$ into a deterministic part
and a noise part,
%
\begin{equation}
  h_i(t) = \langle h_i(t) \rangle + \zeta_i(t).
  \label{eq:h_split}
\end{equation}
%
The deterministic part can be related
to the ``ensemble average'' of $h_i$,
where the ensemble consists of many similar simulation copies
that have experienced the same schedule $\alpha(t)$
and have reached the same $v_i(t)$
at time $t$.
%
The initial states and the stochastic forces
during the process may, however, be different.

@@

To proceed, we first observe that
the average $\bar v = \sum_{i = 1}^n p_i \, v_i$
increases steadily over time:
%
\begin{equation}
\frac{ d \bar v } { d t }
=
\sum_{i = 1}^n p_i \dot v_i
=
\alpha(t) \sum_{i = 1}^n h_i(t) = \alpha(t).
\label{eq:dvbardt}
\end{equation}
%
However, we expect the difference between $v_i$ and $\bar v$,
%
\begin{equation}
  x_i \equiv v_i - \bar v = v_i - \sum_{j = 1}^n p_j \, v_j,
  \label{eq:x_def}
\end{equation}
%
to be small in the asymptotic regime.
Expanding Eq. \eqref{eq:h_ave} in $x_i$ yields
%
\begin{align}
\langle h_i(t) \rangle
&\approx
\frac{                 p_i \, e^{- x_i} }
     { \sum_{ j = 1}^n p_j \, e^{- x_j} }
\approx
\frac{                 p_i (1 - x_i) }
     { \sum_{ j = 1}^n p_j (1 - x_j) }
\notag
\\
&\approx
p_i \, \left(
         1 - x_i + \sum_{j=1}^n p_j \, x_j
       \right)
=
p_i \, (1 - x_i),
\label{eq:hdet_linearize}
\end{align}
%
where we have used $\sum_{j=1}^n p_j = 1$,
and
%
\begin{equation}
  \sum_{i = 1}^n p_i \, x_i = 0,
  \label{eq:px_sum}
\end{equation}
which follows directly from Eq. \eqref{eq:x_def}.
%
From Eqs.
\eqref{eq:vt_diffeq},
\eqref{eq:h_split},
\eqref{eq:dvbardt}-\eqref{eq:hdet_linearize},
we get a set of decoupled equations
%
\begin{equation}
  \dot x_i(t)
  =
  -\alpha(t) \, \left[ x_i(t) - \frac{ \zeta_i(t) } { p_i } \right].
  \label{eq:dxdt_WL}
\end{equation}
\note{The derivation:
$$
\begin{aligned}
  \dot x_i
  &\stackrel{\mathrm{Eq.\; \eqref{eq:x_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \dot v_i - \dot{ \bar v }
  \stackrel{\mathrm{Eqs.\; \eqref{eq:vt_diffeq}\; and\; \eqref{eq:dvbardt}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left( \frac{ h_i } { p_i } - 1 \right)
  \\
  &\stackrel{\mathrm{Eq.\; \eqref{eq:h_split}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left( \frac{ \langle h_i \rangle } { p_i } + \frac{ \zeta_i } { p_i } - 1 \right)
  \stackrel{\mathrm{Eq.\; \eqref{eq:hdet_linearize}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left[ -x_i(t) + \frac{ \zeta_i } { p_i } \right].
\end{aligned}
$$
}

The total error of the bias potential can be written in
terms of $x_i$,
\begin{equation}
\Err
=
\sum_{i = 1}^n p_i \, \langle x_i^2 \rangle
=
\sum_{i = 1}^n p_i \, \left\langle (v_i - \bar v)^2 \right\rangle,
\label{eq:error_sum}
\end{equation}
where
$\bar v = \sum_{i = 1}^n p_i v_i$.
%
Our aim is to find the $\alpha(t)$
that minimizes Eq. \eqref{eq:error_sum}.
%
Fortunately, we find that it suffices to consider
a simpler one-variable problem
in this case.

@@

Consider the following equation
of a single variable:
%
\begin{equation}
\dot x(t) = -\alpha(t) \, \lambda \left[ x(t) - \xi(t) \right],
\label{eq:dxdt_alpha}
\end{equation}
%
where $\xi(t)$ is a generalized noise
that is invariant under time translation:
%
\begin{equation}
\left\langle \xi(t) \, \xi(t') \right\rangle
=
\kappa(t - t').
\notag
%\label{eq:noise_correlation}
\end{equation}
%
Additionally, we require
\begin{equation}
  \lim_{t \rightarrow \pm\infty} \kappa(t) = 0.
  \label{eq:kappa_limit}
\end{equation}
%
For example, for a white noise,
$\kappa(t)$ is proportional to
Dirac's $\delta$-function, $\delta(t)$.
%
We wish to show that the $\alpha(t)$
of minimizing $\langle x^2(t) \rangle$ at long times
is given by
%
\begin{equation}
  \alpha(t) = \frac{1}{\lambda \, (t + t_0)}.
\label{eq:alpha_opt}
\end{equation}

To do so, we first recall
the formal solution of Eq. \eqref{eq:dxdt_alpha}:
%
\begin{equation}
x(T)
=
x(0) \, e^{-\lambda \, q(T)}
+
\int_0^T
  \dot u\bigl( q(t) \bigr) \, \xi(t) \, dt,
\notag
%\label{eq:xt_solution}
\end{equation}
%
where,
%
\begin{equation}
q(t) \equiv \int_0^t \alpha(t') \, dt',
\notag
%\label{eq:q_def}
\end{equation}
%
and
%
\begin{align}
u(q')
&\equiv
e^{-\lambda \, q(T) + \lambda \, q'}.
\label{eq:u_def}
\end{align}
At the end of period $T$, we have
\begin{align}
&\left\langle x^2(T) \right\rangle
=
\langle x^2(0) \,
\rangle e^{-2 \, \lambda \, q(T)}
\notag
\\
&\qquad
+
\int_0^T \int_0^T
  \dot u\bigl( q(t) \bigr) \,
  \dot u\bigl( q(t') \bigr) \,
  \kappa(t - t') \, dt' \, dt.
\label{eq:x2t_average}
\end{align}

>>>>

Free energy calculation\cite{frenkel, newman} is a central theme
in computational physics and chemistry.
%
Given a system,
the task is to compute,
via either Monte Carlo\cite{frenkel, newman, landau_binder} (MC)
and molecular dynamics\cite{frenkel, karplus2002} (MD) simulations,
a distribution, $p^*(z)$,
along a quantity of interest $z$.
%
The negative logarithm,
$-\ln p^*(z)$,
defines a free energy,
or the potential of mean force (PMF).
%
Normally, the distribution $p^*(z)$ is
limited in one or a few narrow $z$ regions.
%
Therefore, to capture the global shape
of the PMF,
one has to run multiple independent simulations
under different umbrella or bias potentials
along $z$\cite{mezei1987, berg1992, lee1993},
and then piece together the entire curve.
%
This solution can be laborious % clumsy in practice
and it often fails to converge quickly
for a complex and glassy system,
because the system can be trapped
in a local free energy minimum.
%for a long time.

@@

We develop the theory
in the following order.
%
In Sec. \ref{sec:background},
we first review the basics of
flat-distribution sampling
and some known aspects on the optimal schedule.
%
This also helps fixing the notations.
%
Then, in Sec. \ref{sec:single-bin},
we illustrate the method of
computing the optimal schedule
on the simplest case of
single-bin (WL) updating scheme.
%
We shall prove the optimality
of the known inverse-time formula.
%
Next, we turn to the general case
of multiple-bin updating schemes
and propose a method of computing
the optimal schedule
in Secs. \ref{sec:multiple-bin}
to \ref{sec:band-matrix}.
%
Finally, we compare different updating schemes
and show that the single-bin scheme
and a class of generalized bandpass schemes
are optimal for convergence
in the long time limit
in Sec. \ref{sec:cmpschemes}.

@@

Consider the problem of computing
the distribution, $p^*_i$,
along a discrete quantity, $i$,
for a given system.
%
%The discreteness of $i$ is natural
%in discrete models.
%
For example, $i$ can be the energy, $E$,
in a lattice spin model; or the temperature index
in a simulated tempering simulation\cite{
marinari1992, lyubartsev1992}.
%
For a continuous quantity, $z$,
which is typical in a molecular system,
we can discretize $z$
such that the integer $i$ represents
the index of a small interval, or a bin,
$(z, z + dz)$.
%
In both cases,
the distribution is normalized as
$\sum_{i = 1}^n p^*_i = 1$.

For a large system,
the distribution $p^*_i$ tends to
be localized,
%
and to find out the global shape
of the PMF, $-\ln p^*_i$,
it is often advantageous to carry out
a biased sampling that targets
a wider distribution, $p_i$.
%
%Here, we refer to simulations that target
%a flat or nearly flat distribution
%as entropic or multicanonical sampling.

To do so, we introduce a bias potential, $V_i$,
which changes the equilibrium distribution to
%
\begin{equation}
  \pi_i \propto p^*_i \, e^{-V_i}.
  \label{eq:pi_p_phi1}
\end{equation}
%
Upon the normalization,
$\sum_{i = 1}^n \pi_i = 1$,
we get
%
\begin{equation}
  \pi_i
  =
  \frac{                p^*_i \, e^{-V_i} }
       { \sum_{j = 1}^n p^*_j \, e^{-V_j} }
  .
  \label{eq:pi_p_phi}
\end{equation}
%
Particularly,
to make $\pi_i$ flat,
the bias potential $V_i$
must coincide with $\ln p^*_i$,
up to an additive constant.
%
In practice, $V_i$ may contain error,
and thus $\pi_i$ need not
coincide with the intended distribution $p_i$.

Here we shall call a simulation that targets
a nearly flat distribution\cite{
dayal2004, trebst2004, singh2011},
$p_i$,
a flat-distribution sampling (FDS) simulation.
%
Roughly speaking, there are two types of FDS methods.
%
The first type,
referred to as the equilibrium FDS, % (E-FDS) methods hereinafter,
uses a fixed bias potential, $V_i$,
(which is estimated beforehand)
during simulation.
%
Examples include the original
entropic or multicanonical sampling\cite{
berg1992, lee1993}
and simulated tempering\cite{
marinari1992, lyubartsev1992}
with $i$ being the index of energy
and that of temperature, respectively.
%
The estimated bias potential, $V_i$,
are often improved
by the histogram, $H_i$, along $i$ accumulated
during simulation.
%
Using $H_i$ as an estimate of $\pi_i$ in
Eq. \eqref{eq:pi_p_phi1}
allows us to correct $V_i$ as
%
\begin{equation}
V^\mathrm{corr}_i
=
V_i
+
\ln \frac{ H_i }
         { p_i }.
\label{eq:vcorr_equil}
\end{equation}
%
\note{This follows from
  $$
  \begin{aligned}
    H_i \approx \pi_i
    &\propto p^*_i \, e^{-V_i},
    \\
    p_i
    &\propto p^*_i \, e^{-V^\mathrm{corr}_i}.
  \end{aligned}
  $$
}
For an unknown system,
it is often difficult
to guess a sufficiently accurate
$V_i$ to start with.
%
Thus,
it can take many iterative rounds of simulations
to achieve the desired flat distribution.
%
\note{Further, a reasonably precise histogram
requires an accumulation period
longer than the autocorrelation time,
which can also be hard to estimate beforehand.}

For this reason, the second type of methods,
referred to as the adaptive FDS methods,
were invented.
%
Examples include
the WL algorithm and metadynamics.
%
These methods update the bias potential
incrementally in regular intervals
and can easily achieve the target distribution.
%
However, they are less well understood theoretically,
as the updating loops
break the microscopic reversibility
of the underlying sampling mechanism.
%
%Thus, for a conservative practitioner,
%the E-FDS methods appear more appealing
%once a sufficiently accurate bias potential is available.
%
To minimize this effect,
many adaptive methods gradually
reduce the updating magnitude
over the course of simulation\cite{
marsili2006,
liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
barducci2008}.
%
Thus, in late stages of a long simulation,
the updating is negligible and
the adaptive simulation tends to
the corresponding equilibrium one.
%
If the bias potential were converged to the exact value,
this particular equilibrium simulation can be thought as
the optimal asymptotic limit of the adaptive counterpart,
delivering the maximal efficiency.

For later convenience,
we introduce a shifted bias potential
%
\begin{equation}
  v_i \equiv V_i - \ln p^*_i + \ln p_i.
  \label{eq:v_def}
\end{equation}
%
Since the last two terms of Eq. \eqref{eq:v_def}
remain constant during the course of simulation,
the updates to $v_i$'s are
the same as the updates to $V_i$'s.
%
In terms of $v_i$'s, Eq. \eqref{eq:pi_p_phi}
becomes
%
\begin{equation}
  \pi_i
  =
  \frac{                p_i \, e^{-v_i} }
       { \sum_{j = 1}^n p_j \, e^{-v_j} }
  \propto
  p_i \, e^{-v_i}.
  \label{eq:pi_p_v}
\end{equation}
%
According to Eq. \eqref{eq:pi_p_phi1},
$v_i$ should ideally be a constant of $i$
such that $\pi_i \approx p_i$.

@@

We shall mainly focus on the adaptive methods below.
%
In the WL algorithm\cite{wang2001, wang2001pre},
the bias potential, $v_i(t)$, is updated
in each MC step $t$ as
%
\begin{equation}
  v_i(t+1)
  =
  v_i(t)
  +
  \delta_{i, \, i(t)}
  \frac{ \alpha(t) } { p_i }
  ,
\label{eq:wl_update}
\end{equation}
%
where $i(t)$ is the bin at step $t$,
and $\alpha(t)$ is the updating magnitude.
%
We refer to this scheme as the \emph{single-bin scheme},
for it applies only to the bias potential
at the current bin, $i(t)$.
%
In a more general \emph{multiple-bin scheme}, however,
several neighboring bins are updated:
%
\begin{equation}
  v_i(t+1)
  =
  v_i(t)
  +
  w_{i, \, i(t)}
  \frac{ \alpha(t) }
       { p_{i(t)} },
  \label{eq:mbin_update}
\end{equation}
%
where $w_{ij}$ are a set of coefficients satisfying
the restrictions to be discussed
in Sec. \ref{sec:updating-matrix}.
%
The Gaussian updating scheme
used in metadynamics
is a multiple-bin scheme.

One can show that with a constant $\alpha(t) = \alpha > 0$,
the distribution collected from
an adaptive FDS simulation
is identical to $p_i$ in long times.
%
However, as the above updates
change $v_i(t)$ continuously,
Eq. \eqref{eq:pi_p_v} no longer holds
at all times,
and the error in $v_i(t)$
depends on the updating magnitude\cite{
  zhou2005, liang2005, laio2005, bussi2006, poulain2006, liang2007,
  morozov2007, zhou2008, morozov2009, crespo2010,
  atchade2011, fort2015}.
%%
%The source of the deviation is two-fold.
%%
%First, since $v_i(t)$ is updated continuously,
%there is a random noise that is proportional
%to $\sqrt \alpha$.
%%
%Second, there is a systematic error
%that comes from the updating dynamics itself,
%as it breaks the Markovian nature
%that underlies Eq. \eqref{eq:pi_p_v}.
%%
%
Thus, one has to decrease $\alpha$ over time
to improve the estimate of the distribution,
$p^*_i$. %via Eq. \eqref{eq:v_def}.

@@

In the original WL papers\cite{
wang2001, wang2001pre},
the updating magnitude $\alpha$
(or $\ln f$ there)
is controlled stage-wise.
%
In each stage, $\alpha$
is kept as a constant,
and the histogram along $i$
is collected and monitored.
%
Once the histogram is sufficiently flat,
we switch to a new stage
with $\alpha$ reduced by a factor of $1/2$\cite{
wang2001, wang2001pre}.
%
This scheme works well for early stages.
%
However, it tends to reduce $\alpha$
too quickly in late stages, making the asymptotic error
saturate\cite{belardinelli2007, belardinelli2007jcp, belardinelli2008}.

%For the single-bin scheme
%used in WL algorithm,
Later,
it was found that
a better way
of controlling the schedule $\alpha(t)$
in late stages
is to follow the formula
%
\begin{equation}
  \alpha(t) = \frac{1}{t},
  \label{eq:alpha_invt}
\end{equation}
%
where $t$,
referred to as the ``time'' below,
is the number of steps
from the beginning of the simulation.
%
This simple yet effective formula has attracted
several studies\cite{belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008, morozov2009,
komura2012, caparica2012, caparica2014}.
%
There are, however, some reservation about
the constant of proportionality:
the optimal formula for $\alpha(t)$
might be $C/t$ with a $C$
different from $1.0$\cite{
morozov2007, zhou2008, morozov2009}.

As Eq. \eqref{eq:alpha_invt} was being developed for the WL algorithm,
several similar ways of
reducing the updating magnitude
were developed for umbrella sampling and metadynamics,
particularly, the self-healing umbrella sampling
and well-tempered metadynamics\cite{
  marsili2006, barducci2008, dickson2011, dama2014}.
%
Our aim below is to find the optimal schedule $\alpha(t)$,
for a general updating scheme.

\subsection{\label{sec:single-bin}
Single-bin updating scheme}

@@

We first approximate Eq. \eqref{eq:wl_update}
by a differential equation
%
\begin{equation}
  \dot v_i(t)
  =
  \alpha(t) \, \frac{ h_i(t) } { p_i },
  \label{eq:vt_diffeq}
\end{equation}
%
where
$\dot v_i(t) \equiv dv_i(t)/dt$,
%
and
\begin{equation}
  h_i(t) = \delta_{i, i(t)}
  ,
  \label{eq:h_def}
\end{equation}
is the instantaneous histogram,
which is equal to $1.0$
for the bin $i(t)$ at time $t$
or zero otherwise.

Next, we split $h_i(t)$ into a deterministic part
and a noise part,
%
\begin{equation}
  h_i(t) = \langle h_i(t) \rangle + \zeta_i(t).
  \label{eq:h_split}
\end{equation}
%
The deterministic part can be related
to the ``ensemble average'' of $h_i$,
where the ensemble consists of many similar simulation copies
that have experienced the same schedule $\alpha(t)$
and have reached the same $v_i(t)$
at time $t$.
%
The initial states and the stochastic forces
during the process may differ, however.

@@

To proceed, we first observe that
the average
$$
\bar v = \sum_{i = 1}^n p_i \, v_i
$$
increases steadily over time:
%
\begin{equation}
\frac{ d \bar v } { d t }
=
\sum_{i = 1}^n p_i \dot v_i
=
\alpha(t) \sum_{i = 1}^n h_i(t) = \alpha(t).
\label{eq:dvbardt}
\end{equation}
%
However, we expect the difference %between $v_i$ and $\bar v$,
%
\begin{equation}
  x_i \equiv v_i - \bar v = v_i - \sum_{j = 1}^n p_j \, v_j,
  \label{eq:x_def}
\end{equation}
%
to be small in the asymptotic regime.
Expanding Eq. \eqref{eq:h_ave} in $x_i$ yields
%
\begin{align}
  \langle h_i(t) \rangle
  &\approx
  \frac{                 p_i \, e^{- x_i} }
       { \sum_{ j = 1}^n p_j \, e^{- x_j} }
  \approx
  p_i \, (1 - x_i),
\label{eq:hdet_linearize}
\end{align}
%
where we have used $\sum_{j=1}^n p_j = 1$,
and
%
\begin{equation}
  \sum_{i = 1}^n p_i \, x_i = 0,
  \label{eq:px_sum}
\end{equation}
which follows directly from Eq. \eqref{eq:x_def}.
\note{For \eqref{eq:hdet_linearize}. Following the second step
$$
\langle h_i(t) \rangle
\approx
\frac{                 p_i (1 - x_i) }
     { \sum_{ j = 1}^n p_j (1 - x_j) }
.
$$
}%
From Eqs.
\eqref{eq:vt_diffeq},
\eqref{eq:h_split},
\eqref{eq:dvbardt}-\eqref{eq:hdet_linearize},
we get a set of decoupled equations
%
\begin{equation}
  \dot x_i(t)
  =
  -\alpha(t) \, \left[ x_i(t) - \frac{ \zeta_i(t) } { p_i } \right].
  \label{eq:dxdt_WL}
\end{equation}
\note{The derivation:
$$
\begin{aligned}
  \dot x_i
  &\stackrel{\mathrm{Eq.\; \eqref{eq:x_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \dot v_i - \dot{ \bar v }
  \stackrel{\mathrm{Eqs.\; \eqref{eq:vt_diffeq}\; and\; \eqref{eq:dvbardt}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left( \frac{ h_i } { p_i } - 1 \right)
  \\
  &\stackrel{\mathrm{Eq.\; \eqref{eq:h_split}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left( \frac{ \langle h_i \rangle } { p_i } + \frac{ \zeta_i } { p_i } - 1 \right)
  \stackrel{\mathrm{Eq.\; \eqref{eq:hdet_linearize}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left[ -x_i(t) + \frac{ \zeta_i } { p_i } \right].
\end{aligned}
$$
}

The total error of the bias potential can be written in
terms of $x_i$,
\begin{equation}
  \Err
  =
  \sum_{i = 1}^n p_i \, \langle x_i^2 \rangle
  =
  \sum_{i = 1}^n p_i \, \left\langle (v_i - \bar v)^2 \right\rangle
  .
\label{eq:error_sum}
\end{equation}
%
Our aim is to find the $\alpha(t)$
that minimizes the error $E$
at the end of the simulation period $T$.
%
Fortunately, it suffices to consider
a simpler one-variable problem.

@@

Consider the following equation
of a single variable:
%
\begin{equation}
\dot x(t) = -\alpha(t) \, \left[ x(t) - \xi(t) \right],
\label{eq:dxdt_alpha}
\end{equation}
%
where $\xi(t)$ is a generalized noise
that is invariant under time translation:
%
\begin{equation}
  \left\langle \xi(t) \, \xi(t') \right\rangle
  =
  \kappa(t - t')
  .
\notag
%\label{eq:noise_correlation}
\end{equation}
%
Additionally,
we assume that $\kappa(t)$ is an even function and
\begin{equation}
  \lim_{t \rightarrow \infty} \kappa(t)
  = 0
  .
  \label{eq:kappa_limit}
\end{equation}
%
For example, for a white noise,
$\kappa(t)$ is proportional to
Dirac's $\delta$-function, $\delta(t)$.
%
We wish to show that the $\alpha(t)$
of minimizing $\langle x^2(t) \rangle$ at long times
is given by
%
\begin{equation}
  \alpha(t) = \frac{ 1 }{ t + t_0 }.
\label{eq:alpha_invt1}
\end{equation}

To do so, we first recall
the formal solution of Eq. \eqref{eq:dxdt_alpha}:
%
\begin{equation}
x(T)
=
x(0) \, e^{-q(T)}
+
\int_0^T
  \dot u\bigl( q(t) \bigr) \, \xi(t) \, dt,
\notag
%\label{eq:xt_solution}
\end{equation}
%
where
%
\begin{equation}
q(t) \equiv \int_0^t \alpha(t') \, dt',
\notag
%\label{eq:q_def}
\end{equation}
%
and
%
\begin{align}
u(q')
&\equiv
e^{q' - q(T)}.
\label{eq:u_def}
\end{align}
At the end of period $T$, we have
\begin{align}
&\left\langle x^2(T) \right\rangle
=
\langle x^2(0) \,
\rangle e^{-2 \, q(T)}
\notag
\\
&\qquad
+
\int_0^T \int_0^T
  \dot u\bigl( q(t) \bigr) \,
  \dot u\bigl( q(t') \bigr) \,
  \kappa(t - t') \, dt' \, dt.
\label{eq:x2t_average}
\end{align}
